\bf The forestecology R package for fitting and assessing neighborhood models of the effect of interspecific competition on the growth of trees

### 2021-03-09 paper/forestecology.tex
3421 -> 3139 via only quick pass of text

\citet{allen_permutation_2020} considers the following model: Let
\(i = 1, \ldots, n_j\) index all \(n_j\) trees of ``focal'' species
group \(j\); let \(j = 1, \ldots, J\) index all \(J\) focal species
groups; and let \(k = 1, \ldots, K\) index all \(K\) ``competitor''
species groups. We model the average annual growth in diameter at breast
height (DBH) \(y_{ij}\) (in centimeters per year) of the \(i^{th}\) tree
of focal species group \(j\) as a linear model \(f\) of the covariates
\(\vec{x}_{ij}\)

@@

In order to evaluate whether competitor species identity matters,
\citet{allen_permutation_2020} run a permutation test where under the
null hypothesis the species identity of all competitors of a focal tree
can be permuted/shuffled:

@@

where the null hypothesis \(H_0\) reflects a hypothesis of no species
grouping-specific effects of competition while the alternative
hypothesis \(H_A\) reflects a hypothesis of species grouping-specific
effects of competition. Furthermore, in order to account for the spatial
autocorrelation inherent to forest data in their estimates of
out-of-sample model error, \citet{allen_permutation_2020} use spatial
cross-validation. Estimates of model error that do not account for this
spatial dependency tend to underestimate the true model error
\citep{roberts_cross-validation_2017}.

We introduce the \texttt{forestecology} R package providing methods and
data for forest ecology model fitting and assessment, available on CRAN
(\url{https://cran.r-project.org/web/packages/forestecology/index.html})
and on GitHub (\url{https://github.com/rudeboybert/forestecology}). The
package implements all aspects of the model in Equation \ref{eq:model}:
model fitting and generating fitted/predicted values, evaluating the
effect of competitor species identity using permutation tests, and
evaluating model performance using spatial cross-validation.

The package designed with ``tidy'' design principles in mind
\citep{wickham_welcome_2019}. Much like many of the \texttt{tidyverse}
component packages, \texttt{forestecology} is designed with verb-named
functions that can be modularly composed in sequence using the pipe
\texttt{\%\textgreater{}\%} operator \citep{bache_pipe_2020}. As we
articulate in Section \ref{casestudy}, these functions delineate the key
steps in our analysis sequence. Furthermore, the inputs and outputs of
nearly all of our functions use the same ``simple features for R'' data
structures as implemented in the \texttt{sf} package for standardized
and \texttt{tidyverse}-friendly support for spatial vector data
\citep{pebesma_simple_2018}

Currently the package only implements the Bayesian linear regression
model of tree growth based on neighborhood competition detailed in
Equation \ref{eq:model}. As we demonstrate in Section
\ref{model-fit-predict} however, the fitting of this model is
self-contained in a single function \texttt{comp\_bayes\_lm()}. This
function returns an object of S3 class type \texttt{comp\_bayes\_lm}
with generic methods implemented to print, make predictions using, and
plot all results. Therefore the package can be modularly extended to fit
other models as long as they are coded into a function similar to
\texttt{comp\_bayes\_lm()} and has equivalent generic methods
implemented.

@@

We present a case-study of the \texttt{forestecology} package's
functionality on data from the Smithsonian Conservation Biology
Institute (SCBI) large forest dynamics plot in Front Royal, VA, USA,
which is part of the ForestGEO global network of research sites
\citep[\citet{andersonteixeira_ctfs-forestgeo_2015}]{bourg_initial_2013}
\citep{bourg_initial_2013}. The 25.6 ha (640 x 400 m) plot is located at
the intersection of three of the major physiographic provinces of the
eastern US---the Blue Ridge, Ridge and Valley, and Piedmont
provinces---and is adjacent to the northern end of Shenandoah National
Park.

The \texttt{forestecology} package has the following ecological goals:
1) to evaluate the effect of competitor species identity using
permutation tests and 2) to evaluate model performance using spatial
cross-validation. To achieve these goals, we outline a basic analysis
sequence comprising of these four main steps:

@@

  \begin{enumerate}
  \def\labelenumii{\arabic{enumii}.}
  \tightlist
  \item
    Fit model.
  \item
    Compute fitted/predicted values.
  \item
    Visualize posterior distributions.
  \end{enumerate}
\end{enumerate}

@@

The first step is to compute the growth of trees using data from two
censuses. \texttt{compute\_growth()} computes average annual growth
assuming census data that roughly follows ForestGEO standards. Despite
such standards, minor variations will still exist between sites, thereby
necessitating some data wrangling and checking. For example, the SCBI
site records all diameters at breast height (DBH) in millimeters
\citep{bourg_initial_2013}, whereas the Michigan Big Woods site records
them in centimeters \citep{allen_michigan_2020}.

We first load both 2008 and 2014 SCBI census data \texttt{.csv} files as
they existed on GitHub on November 20, 2020
\citep{gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020} and perform
some data wrangling to both data sets. We then only consider a 9 ha
subsection of the 25.6 ha of the SCBI site, \texttt{gx} from 0--300
instead of 0--400 and \texttt{gy} from 300--600 instead of 0--640, in
order to speed up computation for purposes of this example.

@@

These two data frames are then supplied as arguments to
\texttt{compute\_growth()}, along with the \texttt{id} argument that
specifies the variable that uniquely identifies each tree-stem. Note
furthermore that we discard all resprouts in the later census (those
with \texttt{code\ ==\ R}), since we are only interested in the diameter
growth of surviving, and not resprouted, stems.

@@

The output \texttt{growth\_scbi} is a single data frame of class
\texttt{sf} that includes variables \texttt{growth}, the average annual
growth in DBH (cm \(\cdot\) y\textsuperscript{-1}) for all stems that
were alive at both time points, and \texttt{geometry}, the \texttt{sf}
package's encoding of geolocations of type
\texttt{\textless{}POINT\textgreater{}}. In addition the species
variable \texttt{sp} is returned as a factor.\footnote{In our spatial
  cross-validation algorithm in Section \ref{spatial-cross-validation}
  issues can occur when rare species do not occur in the training set,
  but then are encountered in the test set. This risk is mitigated by
  representing \texttt{sp} as a factor variable, which has a complete
  list of all levels of the categorical variable.} Given that
\texttt{growth\_scbi} is of class \texttt{sf}, it can be easily plotted
in \texttt{ggplot2} using the \texttt{geom\_sf()} geometry as seen in
Figure \ref{fig:scbi-trees}.

@@

The next step is to add additional spatial information to
\texttt{growth\_scbi}. The first element we add is a ``buffer region''
to the periphery of the study region. Since some of our model's
explanatory variables are cumulative, we must ensure that all trees
being modeled are not biased to have different neighbor structures. This
is of concern for trees at the boundary of study regions, for which all
neighbors will not be included in the censused stems. In order to
account for such edge effects, only trees that are not part of this
buffer region, i.e.~are part of the interior of the study region, will
have their growth modeled \citep{waller_applied_2004}.

Our model of interspecific competition relies on a spatial definition of
who the competitor trees are for focal trees of interest: all trees
within a distance \texttt{comp\_dist} of a focal tree are considered its
competitors. In our case we set this value at 7.5m, a value informed by
other studies \citep[\citet{uriarte_spatially_2004},
\citet{canham_neighborhood_2006}]{canham_neighborhood_2004}. Using this
value along with a manually constructed \texttt{sf} object
representation of the study region's boundary, we apply the
\texttt{add\_buffer\_variable()} to \texttt{growth\_scbi} to add a
\texttt{buffer} boolean variable. All trees with \texttt{buffer} as
\texttt{FALSE} will be our focal trees whose growth will be modeled,
whereas those with \texttt{TRUE} will only be considered as competitor
trees.

@@

The second element of spatial information are blocks corresponding to
folds of a spatial cross-validation algorithm used to estimate
out-of-sample model error. Conventional cross-validation algorithms
assign observations to folds by randomly resampling individual
observations. However, many of these algorithms assume that the
observations are independent. In the case of forest census data,
observations exhibit spatial autocorrelation. We therefore incorporate
this spatial dependence into the cross-validation algorithm with our
spatial blocks of trees \citep[
\citet{pohjankukka_estimating_2017}]{roberts_cross-validation_2017}.

In the example below, we first manually define four folds that partition
the study region as an \texttt{sf} object. We then use the output of the
\texttt{spatialBlock()} function from the \texttt{blockCV} package to
associate each tree in \texttt{growth\_scbi} to the correct fold
\texttt{foldID} \citep{valavi_blockcv_2019}. \footnote{In the Supporting
  Information we present an example where the folds themselves are also
  created automatically using \texttt{spatialBlock()} given a specified
  \texttt{cv\_block\_size}, as opposed to manually as in the example.}
Figure \ref{fig:scbi-spatial-information} illustrates the net effect of
adding these two elements of spatial information to
\texttt{growth\_scbi}.

@@

The next step is to identify all focal trees and their corresponding
competitor trees. More specifically, identify all trees that are not
part of the buffer region, have a valid \texttt{growth} measurement, and
have at least one neighbor within 7.5m.
\texttt{create\_focal\_vs\_comp()} returns a new data frame of type
\texttt{sf}. On top of previously detailed arguments \texttt{comp\_dist}
and \texttt{id}, \texttt{create\_focal\_vs\_comp()} also requires an
\texttt{sf} object representation of the spatial cross-validation
blocks/folds as seen in Section \ref{spatial-information}.

@@

The resulting \texttt{focal\_vs\_comp\_scbi} has 6296 rows, representing
the subset of the 7954 trees in \texttt{growth\_scbi} that will be
considered as focal trees. Two new variables \texttt{focal\_ID} and
\texttt{focal\_sp} relate to tree-stem identification and species
information. Most notably however is a new variable \texttt{comp} which
contains information on all competitor trees saved in \texttt{tidyr}
package list-column format \citep{tidyr_package}. We flatten the
\texttt{comp} list-column for the tree with \texttt{focal\_ID} 4 in the
first row, here a \texttt{tibble\ {[}20\ ×\ 4{]}}, into regular columns
using \texttt{unnest()} from the \texttt{tidyr} package.

@@

We observe that for this focal tree, we have 4 variables of information
on its 20 competitor trees: their unique tree-stem ID number, their
distance to the focal tree (all \(\leq\) 7.5), their species, and their
basal area (in m\(^2\)) calculated as
\(\frac{\pi \times (\text{DBH/2})^2}{10000}\) where \(DBH\) is the value
from the earlier of the two censuses in cm. Saving our focal versus
competitor information in list-column minimizes redundancy since we do
not repeat information on the focal tree 20 times. The spatial
distribution of these trees is visualized in Figure
\ref{fig:scbi-focal-vs-comp-map}.

@@

The final step is to fit a model for the growth of all focal trees. We
fit the competition Bayesian linear regression model outlined in
Equation \ref{eq:model} using \texttt{comp\_bayes\_lm()}, which has an
option to specify prior distributions on all parameters of interest
(chosen to be the defaults specified in \texttt{?comp\_bayes\_lm}).

@@

The returned \texttt{comp\_bayes\_lm\_scbi} output is an object of S3
class type \texttt{comp\_bayes\_lm} containing the posterior values of
all parameters in our competition Bayesian linear regression. This class
of object includes three generic methods. First, the generic for
\texttt{print()} displays the names of all prior \& posterior parameters
along with the model formula:

@@

Next, the generic for \texttt{predict()} takes the posterior parameter
values in \texttt{comp\_bayes\_lm\_scbi} and the predictor variables in
\texttt{newdata} and outputs a vector \texttt{growth\_hat} of
fitted/predicted values \(\widehat{y}\) of the DBH for each focal tree
computed from the posterior predictive distribution.

@@

Lastly, the generic for \texttt{ggplot2::autoplot()} allows us to plot
the posterior distribution of all parameters in Figure
\ref{fig:scbi-posterior-viz}.

@@

\CommentTok{# Combine plots using patchwork}
\NormalTok{(plot1 }\OperatorTok{|}\StringTok{ }\NormalTok{plot2) }\OperatorTok{/}\StringTok{ }\NormalTok{plot3}
\end{Highlighting}
\end{Shaded}

@@

These plots visualize the posterior distributions of parameters from
Equation \ref{eq:model}. For many package users they will be of interest
because they give insight into the species-specific competitive
interactions. Setting \texttt{type\ =\ "intercepts"} returns
species-specific posterior distributions for \(\beta_{0,j}\) and
\texttt{type\ =\ "dbh\_slopes"} for \(\beta_{dbh,j}\). Setting
\texttt{type\ =\ "competition"} returns competition coefficients
\(\lambda_{j,k}\) where negative values indicate a competitor species
which slows the growth of a focal species. Here, for example, we see
that \texttt{litu} tulip poplars have a strong negative effect on the
growth of conspecifics but relatively lesser effect on neighbors of the
other two species.

Currently the \texttt{forestecology} package can only fit the
competition Bayesian linear regression model outlined in Equation
\ref{eq:model}. However, it can be extended to any model implemented in
a function similar to \texttt{comp\_bayes\_lm()} that uses data frames
of similar format to \texttt{focal\_vs\_comp} as input.

\hypertarget{evaluate-the-effect-of-competitor-species-identity-using-permutation-tests}{%
\subsection{Evaluate the effect of competitor species identity using
permutation
tests}\label{evaluate-the-effect-of-competitor-species-identity-using-permutation-tests}}

To evaluate the effect of competitor species identity, we use the four
steps of our analysis sequence answer along with a permutation test:
Under a null hypothesis where competitor species identity does not
matter, we permute/shuffle this variable within each focal tree, compute
the RMSE (the test statistic of interest), repeat this process several
times to construct a null distribution of the RMSE, and compare it to
the observed RMSE to assess significance. Going back to our example in
Section \ref{focal-vs-comp} of focal tree with \texttt{focal\_ID} 4 and
its 20 competitors, the permutation test randomly resamples only the
\texttt{comp\_sp} variable with replacement, leaving all other variables
intact. The resampling with replacement is nested within each focal tree
in order to preserve neighborhood structure. We once again use
\texttt{comp\_bayes\_lm()} as in Section \ref{model-fit-predict}, but
with \texttt{run\_shuffle\ =\ TRUE}.

@@

The resulting RMSE of 0.131 based on the permutation test is larger than
the earlier RMSE of 0.128, suggesting that models that do incorporate
competitor species identity better fit the data.

@@

We answer the second of our two questions: how can we obtain an accurate
estimate of model performance/error? The model fits and predictions in
Section \ref{model-fit-predict} all suffer from a common failing: they
use the same data to both fit the model and to assess the model's
performance using the RMSE. As argued by
\citet{roberts_cross-validation_2017}, this can lead to overly
optimistic assessments of model quality as the models can be overfit, in
particular in situations where spatial-autocorrelation is present. To
mitigate the effects of such overfitting, we use a spatially block
cross-validation algorithm.

To this end, we use the \texttt{foldID} variable defined in Section
\ref{spatial-information} whereby all focal trees are assigned to one of
4 spatially contiguous blocks that act as folds in our cross-validation
routine. Figure \ref{fig:scbi-spatial-cross-validation-schematic}
presents a schematic illustrating this scheme. We fit the model to all
focal trees in the training set, apply the model to all focal trees in
the test set to compute fitted/predicted values, and compute the RMSE of
the observed versus predicted growths. We repeat this procedure 3 more
times with each of the three remaining folds acting as the test set and
then average all four resulting RMSE's. Furthermore, in order to
maintain spatial independence between the test and training set, a
``fold buffer'' that extend outwards from the boundary of the test set
is computed; all trees falling within this fold buffer are excluded from
the training set.

@@

\caption{Schematic of spatial cross-validation: Using the k = 1 fold (bottom-left) as the test set, k = 2 through 4 as the training set, along with a "fold buffer."}\label{fig:scbi-spatial-cross-validation-schematic}
\end{figure}

This algorithm is implemented in \texttt{run\_cv()}, which is a wrapper
function to both \texttt{comp\_bayes\_lm()} that fits the model and
\texttt{predict()} that returns fitted/predicted values. We once again
compute the RMSE.

@@

The resulting RMSE of 0.14 computed using cross-validation is larger
than the earlier RMSE of 0.128, suggesting that models that do not take
the inherent spatial autocorrelation of the data into account generate
error estimates that are overly optimistic; in our case RMSE's that are
too low.

@@

\texttt{run\_cv()} also accepts the \texttt{run\_shuffle} argument. This
permutes the competitor species, as described above, but does so when
calculating predicted growth with the cross validated scheme. Figure
\ref{fig:scbi-simulation} compares model performance when permuting
competitor species and calculating RMSE with and without
cross-validation. Without cross-validation the competitor identity did
matter, the non-permuted competitor species had a much lower RMSE than
the permuted one. But once we include the spatial cross-validation, this
improvement disappears. These results suggest that in this 9 ha subplot
of the SCBI plot competitive interactions do not depend on the identity
of the competitor, which is the opposite of what has been observed in
other locations (\citet{allen_permutation_2020}
\citet{uriarte_spatially_2004}). This highlights the importance of
cross-validation, without it the model was overfit.

@@

The \texttt{forestecology} package provides an accessible way to fit and
test models of neighborhood competition. While the package is designed
with ForestGEO plot data in mind, we envision that it can be modified to
work on i) any single large, mapped forest plot in which at least two
measurements of each individual have been taken, e.g.~the US Forest
Service Forest Inventory and Analysis plots or ii) more generally to
model interactions of any community of mapped sessile organisms
\citep{smith_forest_2002}. In future versions of \texttt{forestecology}
we also hope to make it possible to model plant mortality in addition to
plant growth. The package follows the guidelines for \texttt{tidy} data,
leverages the \texttt{sf} package for spatial data, and S3 open-oriented
model structure. We hope that the package will increase the use of
neighborhood competition models to better understand what structures
plant competition.

>>>>

\citet{allen_permutation_2020} considers the following model: Let
\(i = 1, \ldots, n_j\) index all \(n_j\) trees of ``focal'' species
group \(j\); let \(j = 1, \ldots, J\) index all \(J\) focal species
groups; and let \(k = 1, \ldots, K\) index all \(K\) ``competitor''
species groups. They use the following linear model the average annual
growth in diameter at breast height (DBH) \(y_{ij}\) (in
centimeters/year) of the \(i^{th}\) tree of focal species group \(j\):

@@

In order to evaluate whether competitor species identity matters,
\citet{allen_permutation_2020} run a permutation test where a null
hypothesis of no species grouping-specific effects of competition is
assumed, and thus the species identity of all competitors can be
permuted/shuffled:

@@

Furthermore, in order to account for the spatial autocorrelation
inherent to forest data in their estimates of out-of-sample model error,
\citet{allen_permutation_2020} use spatial cross-validation. Estimates
of model error that do not account for this spatial dependency tend to
underestimate the true model error
\citep{roberts_cross-validation_2017}.

We introduce the \texttt{forestecology} R package providing methods and
data for forest ecology model fitting and assessment, available on CRAN
(\url{https://cran.r-project.org/web/packages/forestecology/index.html})
and on GitHub (\url{https://github.com/rudeboybert/forestecology}). The
package implements all aspects of the model in Equation \ref{eq:model}:
model fitting and generating predicted values, evaluating the effect of
competitor species identity using permutation tests, and evaluating
model performance using spatial cross-validation.

The package designed with ``tidy'' design principles in mind
\citep{wickham_welcome_2019}. Much like all \texttt{tidyverse} packages,
\texttt{forestecology} is has verb-named functions that can be modularly
composed using the pipe \texttt{\%\textgreater{}\%} operator to complete
all the necessary steps in the analysis sequence
\citep{bache_pipe_2020}. Furthermore, the inputs and outputs of most of
our functions use the same ``simple features for R'' data structures
from the \texttt{sf} package for standardized and
\texttt{tidyverse}-friendly wrangling and visualizing of spatial data
\citep{pebesma_simple_2018}

Currently the package only implements the Bayesian linear regression
model detailed in Equation \ref{eq:model}. As we demonstrate in Section
\ref{model-fit-predict} however, the fitting of this model is
self-contained in a single function \texttt{comp\_bayes\_lm()} which
returns an object of S3 class type \texttt{comp\_bayes\_lm}. This class
has generic methods implemented to print, make predictions using, and
plot all results. Therefore the package can be modularly extended to fit
other models as long as they are coded similarly as
\texttt{comp\_bayes\_lm()} and have equivalent generic methods
implemented.

@@

We present a case-study of the \texttt{forestecology} package's
functionality on data from the Smithsonian Conservation Biology
Institute (SCBI) large forest dynamics plot in Front Royal, VA, USA,
part of the ForestGEO global network of research sites
\citep[\citet{andersonteixeira_ctfs-forestgeo_2015}]{bourg_initial_2013}
\citep{bourg_initial_2013}. The 25.6 ha (640 x 400 m) plot is located at
the intersection of three of the major physiographic provinces of the
eastern US---the Blue Ridge, Ridge and Valley, and Piedmont
provinces---and is adjacent to the northern end of Shenandoah National
Park.

The package has the following ecological goals: to evaluate i) the
effect of competitor species identity using permutation tests and ii)
model performance using spatial cross-validation. We outline a basic
analysis sequence comprising of the following four main steps:

@@

  \begin{enumerate}
  \def\labelenumii{\arabic{enumii}.}
  \tightlist
  \item
    Fit model.
  \item
    Compute predicted values.
  \item
    Visualize posterior distributions.
  \end{enumerate}
\end{enumerate}

@@

The first step is to compute the growth of trees using data from two
censuses. \texttt{compute\_growth()} computes the average annual growth
based on census data that roughly follows ForestGEO standards. Despite
such standards, minor variations will still exist between sites, thereby
necessitating some data wrangling. For example, the SCBI site records
all DBH values in millimeters \citep{bourg_initial_2013}, whereas the
Michigan Big Woods site records them in centimeters
\citep{allen_michigan_2020}.

We first load both 2008 and 2014 SCBI census \texttt{.csv} files as they
existed on GitHub on 2020/11/20 and perform some data wrangling to both
data sets \citep{gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020}.
We then only consider a 9 ha subsection of the 25.6 ha of the SCBI site
in order to speed up computation for this example: \texttt{gx} from
0--300 instead of 0--400 and \texttt{gy} from 300--600 instead of
0--640.

@@

These two data frames are then used as inputs to
\texttt{compute\_growth()}, along with the \texttt{id} argument
specifying the variable that uniquely identifies each tree-stem. Note
that we also discard all resprouts in the later census with
\texttt{code\ ==\ R}, since we are only interested in the growth of
surviving, and not resprouted, stems.

@@

The output \texttt{growth\_scbi} is a data frame of class \texttt{sf}
that includes variables the average annual \texttt{growth} in DBH (cm
\(\cdot\) y\textsuperscript{-1}) for all stems that were alive at both
time points, the \texttt{sf} package's encoding of geolocations of
\texttt{geometry} type \texttt{\textless{}POINT\textgreater{}}, and the
species variable \texttt{sp} converted to a factor. Given that
\texttt{growth\_scbi} is of class \texttt{sf}, it can be easily plotted
in \texttt{ggplot2} using \texttt{geom\_sf()} as seen in Figure
\ref{fig:scbi-trees}.

@@

The next step is to add spatial information to \texttt{growth\_scbi}.
The first element we add is a ``buffer region'' to the periphery of the
study region. Since some of our model's explanatory variables are
cumulative, we must ensure that all trees being modeled are not biased
to have different neighbor structures. This is of concern for trees at
the boundary of the study region whose neighbors will not all be
included in the censused stems. To account for such edge effects only
trees that are not part of this buffer region, i.e.~are part of the
interior of the study region, will have their growth modeled
\citep{waller_applied_2004}.

Our model of interspecific competition relies on a spatial definition of
who competitor trees are: all trees within a distance
\texttt{comp\_dist} of a focal tree. In our case we set
\texttt{comp\_dist} to 7.5m, a value informed by other studies
\citep[\citet{uriarte_spatially_2004},
\citet{canham_neighborhood_2006}]{canham_neighborhood_2004}. We use
\texttt{comp\_dist} and a manually constructed \texttt{sf}
representation of the study region's boundary as inputs to
\texttt{add\_buffer\_variable()} to add a \texttt{buffer} boolean
variable to \texttt{growth\_scbi}. All trees with \texttt{buffer} equal
to \texttt{FALSE} will be our focal trees whose growth will be modeled,
whereas those with \texttt{TRUE} will only act as competitor trees.

@@

The second element of spatial information we add are blocks
corresponding to folds of a spatial cross-validation algorithm.
Conventional cross-validation algorithms assign observations to folds by
randomly resampling individual observations that are assumed
independent. In the case of forest census data however, observations
exhibit spatial autocorrelation. We therefore incorporate this
dependence into the cross-validation algorithm by resampling spatial
blocks of trees \citep[
\citet{pohjankukka_estimating_2017}]{roberts_cross-validation_2017}.

We first manually an \texttt{sf} object defining four folds that
partition the study region. We then use the output of the
\texttt{spatialBlock()} function from the \texttt{blockCV} package to
associate each tree in \texttt{growth\_scbi} to the correct
\texttt{foldID} \citep{valavi_blockcv_2019}.\footnote{In the Supporting
  Information we present an example where the folds themselves are
  created automatically, as opposed to manually as in the example.} This
\texttt{foldID} variable will be used in Section
\ref{spatial-cross-validation}.

Figure \ref{fig:scbi-spatial-information} illustrates the net effect of
adding these two elements of spatial information to
\texttt{growth\_scbi}.

@@

The next step is to identify all focal trees and their corresponding
competitor trees. More specifically, identify all trees that are not
part of the buffer region, have a valid \texttt{growth} measurement, and
have at least one neighbor within 7.5m. We do this using
\texttt{create\_focal\_vs\_comp()}, which takes the previously detailed
arguments \texttt{comp\_dist} and \texttt{id} as well as the \texttt{sf}
representation of the spatial cross-validation blocks and returns a new
data frame \texttt{focal\_vs\_comp\_scbi}.

@@

The resulting \texttt{focal\_vs\_comp\_scbi} has 6296 rows, representing
the subset of the 7954 trees in \texttt{growth\_scbi} that will be
considered as focal trees. The variables \texttt{focal\_ID} and
\texttt{focal\_sp} relate to tree-stem identification and species
information. Most notably however is the variable \texttt{comp}, which
contains information on all competitor trees saved in \texttt{tidyr}
package list-column format \citep{tidyr_package}. To inspect this
information, we flatten the \texttt{comp} list-column for the tree with
\texttt{focal\_ID} 4 in the first row, here a
\texttt{tibble\ {[}20\ ×\ 4{]}}, into regular columns using
\texttt{unnest()} from the \texttt{tidyr} package.

@@

We observe 4 variables describing 20 competitor trees: their unique
tree-stem ID, their distance to the focal tree (all \(\leq\) 7.5), their
species, and their basal area (in m\(^2\)) calculated as
\(\frac{\pi \times (\text{DBH/2})^2}{10000}\) for \(DBH\) in cm from the
earlier census. Saving competitor information in list-column format
minimizes redundancy since we do not need to repeat information on the
focal tree 20 times. We visualize the spatial distribution of these
trees in Figure \ref{fig:scbi-focal-vs-comp-map}.

@@

The final step is to fit the competition Bayesian linear regression
model for tree growth outlined in Equation \ref{eq:model} using
\texttt{comp\_bayes\_lm()}. This function has an option to specify prior
distributions on all parameters of interest, chosen here to be the
defaults detailed in \texttt{?comp\_bayes\_lm}.

@@

The resutling \texttt{comp\_bayes\_lm\_scbi} is an object of S3 class
type \texttt{comp\_bayes\_lm} containing the posterior values of all
parameters of our model. Furthermore, this class of object includes
generics for three methods. First, the generic for \texttt{print()}
displays the names of all prior and posterior parameters along with the
model formula:

@@

Next, the generic for \texttt{predict()} takes the posterior parameter
values in \texttt{comp\_bayes\_lm\_scbi} and a \texttt{newdata} data
frame and outputs a vector \texttt{growth\_hat} of predicted values
\(\widehat{y}\) of the DBH for each focal tree computed from the
posterior predictive distribution.

@@

Lastly, the generic for \texttt{ggplot2::autoplot()} allows us to
visualize the posterior distribution of all parameters, as seen in
Figure \ref{fig:scbi-posterior-viz}. Setting \texttt{type} to
\texttt{"intercepts"} and \texttt{"dbh\_slopes"} returns
species-specific posterior distributions for \(\beta_{0,j}\) and
\(\beta_{dbh,j}\) respectively, while setting
\texttt{type\ =\ "competition"} returns competition coefficients
\(\lambda_{j,k}\).

@@

\CommentTok{# Combine plots using the patchwork package}
\NormalTok{(plot1 }\OperatorTok{|}\StringTok{ }\NormalTok{plot2) }\OperatorTok{/}\StringTok{ }\NormalTok{plot3}
\end{Highlighting}
\end{Shaded}

@@

For many users the visualizations relating to \(\lambda_{j,k}\) will be
of particular interest as they provide insight into species competitive
interactions where negative values indicate a competitor species which
slows the growth of a focal species. Here, for example, we see that
tulip poplars (\texttt{litu}) have a strong negative effect on the
growth of conspecifics but relatively lesser effect on neighbors of the
other two species.

Currently the \texttt{forestecology} package can only fit the
competition Bayesian linear regression model outlined in Equation
\ref{eq:model}. However, it can be extended to any model as long as it
is implemented in a function similar to \texttt{comp\_bayes\_lm()}.

\hypertarget{permutation-test}{%
\subsection{Evaluate the effect of competitor species identity using
permutation tests}\label{permutation-test}}

To evaluate the effect of competitor species identity, we use the above
four steps along with a permutation test. Under a null hypothesis where
competitor species identity does not matter, we permute the competitor
species identity within each focal tree, compute the RMSE test
statistic, repeat this process several times to construct a null
distribution, and finally compare it to the observed RMSE to assess
significance. Going back to our example in Section \ref{focal-vs-comp}
of focal tree with \texttt{focal\_ID} 4 and its 20 competitors, the
permutation test randomly resamples only the \texttt{comp\_sp} variable
without replacement, leaving all other variables intact. The resampling
without replacement is nested within each focal tree in order to
preserve neighborhood structure. We once again use
\texttt{comp\_bayes\_lm()}, but this time setting
\texttt{run\_shuffle\ =\ TRUE}.

@@

The resulting permutation test RMSE of 0.131 is larger than the earlier
RMSE of 0.128, suggesting that models that do incorporate competitor
species identity better fit the data.

@@

To evaluate model performance, we use spatial cross-validation. The
model fits and predictions in Section \ref{model-fit-predict} use the
same data to both fit the model and to assess the model's performance.
Given the inherent spatial-autocorrelation of our data, this can
potentially lead to potentially overfit models
\citep{roberts_cross-validation_2017}. To mitigate this potential
overfitting, we use the spatial cross-validation blocking scheme encoded
in the \texttt{foldID} variable from Section \ref{spatial-information}
and visualized in Figure \ref{fig:scbi-spatial-information}.

At each iteration of our cross-validation, one fold will act as the test
set with the remaining three acting as the training set. We fit the
model to all focal trees in the training set, apply the model to all
focal trees in the test set, compute predicted values, and compute the
RMSE. Furthermore, to maintain spatial independence between the test and
training set, a ``fold buffer'' that extend outwards from the boundary
of the test set is computed; all trees falling within this fold buffer
are excluded from the training set (see Figure
\ref{fig:scbi-spatial-cross-validation-schematic}).

@@

\caption{Schematic of spatial cross-validation: Using the k = 1 fold (bottom-left) as the test set, k = 2 through 4 as the training set, along with a fold buffer.}\label{fig:scbi-spatial-cross-validation-schematic}
\end{figure}

This is repeated for each of the four folds acting as the test set and
then the four resulting RMSE's are averaged to provide a single estimate
of model error. This algorithm is implemented in \texttt{run\_cv()},
which acts as a wrapper function to both \texttt{comp\_bayes\_lm()} that
fits the model and \texttt{predict()} that returns predicted values.

@@

The resulting RMSE of 0.14 computed using cross-validation is larger
than the earlier RMSE of 0.128, suggesting that models that do not take
spatial autocorrelation account generate moel error estimates that are
overly optimistic. In our case, RMSE values that are too low.

@@

\texttt{run\_cv()} also accepts the \texttt{run\_shuffle} argument in
order to permute competitor species identity as described in Section
\ref{permutation-test}. Figure \ref{fig:scbi-simulation} compares model
performance for 49 permutations of competitor species and calculating
RMSE, both with and without cross-validation. Without cross-validation,
competitor species identity did matter as the observed RMSE was lower
than the permutation null distribution of RMSE. However, once we
incorporate spatial cross-validation, this improvement disappears. These
results suggest that in this 9 ha subplot of the SCBI plot competitive
interactions do not depend on the identity of the competitor, which is
the opposite of what has been observed in other locations \citep[
\citet{uriarte_spatially_2004}]{allen_permutation_2020}. Furthermore,
the larger cross-validated RMSE's are indicative of the importance of
being vigilant against model overfitting.

@@

The \texttt{forestecology} package provides an accessible way to fit and
test models of neighborhood competition. While the package is designed
with ForestGEO plot data in mind, we envision that it can be modified to
work on i) any single large, mapped forest plot in which at least two
measurements of each individual have been taken, e.g.~the US Forest
Service Forest Inventory and Analysis plots, or ii) more generally to
model interactions of any community of mapped sessile organisms
\citep{smith_forest_2002}. In future versions of \texttt{forestecology}
we also hope to include models that account for tree mortality in
addition to tree growth. The package follows the guidelines for
\texttt{tidy} data, leverages the \texttt{sf} package for spatial data,
and S3 open-oriented model structure. We hope that the package will
increase the use of neighborhood competition models to better understand
what structures plant competition.
